---
title: 'Ecce Signum: Illustration'
output:
  html_document:
    df_print: paged
---

# New Zealand Arrivals

This illustration considers a daily immigration series (**imm** for short): New 
Zealand residents arriving in New Zealand after an absence of less than 12 
months, covering the period January 1, 2008 through July 31, 2012. The series 
consists of New Zealand residents arriving in New Zealand after an absence of 
less than 12 months. These are public use data produced by Statistics New 
Zealand via a customized extract, and correspond to a portion of the 
"daily border crossings - arrivals" tab (Total) of the Travel category in the 
Covid-19 portal: https://www.stats.govt.nz/experimental/covid-19-data-portal.
  
## Loading Packages 

This code installs and loads the packages, and loads the data into a variable 
named **imm**.
   	
```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(devtools)
library(Rcpp)
devtools::install_github("tuckermcelroy/sigex")
library(sigex)
```
  
- We do some initial processing.
- The original data is grouped into "months" of 31 days, with extra days of 
value zero added for months of shorter length; our code removes these features, 
as well as a monthly summation.

```{r}
n.months <- dim(imm)[1]/32
imm <- imm[-seq(1,n.months)*32,]
imm <- matrix(imm[imm != 0],ncol=6) 
```

- Also we load a set of regressors that will be used in later analysis.
- The first column is a regressor that captures a moving holiday effect for 
Easter, and the next six columns correspond to school holidays for public 
schools in New Zealand: an effect is estimated for the start and end of the 
holiday for the first three school vacations in each year. The last three 
columns correspond to Y2K effects (but will not be used in this script).
 
```{r}
nz.path <- system.file('extdata','NZregressors.txt',package='sigex')
NZregs <- read.table(nz.path)
```
  
## Enter Metadata

- We perform an analysis of one series by using an embedding to the weekly
frequency.
- First enter metadata, including starting date and frequency.
   	 
```{r}
start.date <- c(9,1,1997)
end.date <- day2date(dim(imm)[1]-1,start.date)
period <- 365
```
    
- Next we do some calendar calculations.

```{r}
start.day <- date2day(start.date[1],start.date[2],start.date[3])
end.day <- date2day(end.date[1],end.date[2],end.date[3])
begin <- c(start.date[3],start.day)
end <- c(end.date[3],end.day)
```

- Finally, there is a call to **sigex.load**, which associates labels to the 
data.
- The last argument of **sigex.load** generates a time series plot.

```{r}
dataALL.ts <- sigex.load(imm,begin,period,
                  c("NZArr","NZDep","VisArr","VisDep","PLTArr","PLTDep"),TRUE)
```
   
## Select Spans and Transforms

- On the basis of the above plot, we decide to utilize a log transformation for 
the data.
- The *aggregate* option will sum across the indicated series, in case we want 
to analyze an aggregate. Here we set it to FALSE.
- We can also select a subcomponent of series with *subseries*. Here we focus 
on the first series only.
- The *range* argument of **sigex.prep** will select a subset of dates for the
time series to be analyzed.
   
```{r}   
transform <- "log"
aggregate <- FALSE
subseries <- 1
range <- list(c(2008,1),end)
dataONE.ts <- sigex.prep(dataALL.ts,transform,aggregate,subseries,range,TRUE)
```
   
## Spectral Exploratory Analysis

- In order to get an idea about model specification, we can examine spectral 
estimates, using **sigex.specar**.
- We can look at raw data or differenced data (growth rates).

```{r}
par(mfrow=c(1,1))
for(i in 1:length(subseries)) { sigex.specar(dataONE.ts,FALSE,i,7) }
```

```{r}
par(mfrow=c(1,1))
for(i in 1:length(subseries)) { sigex.specar(dataONE.ts,TRUE,i,7) }
```

## Embed to Weekly 

- We set the first day of the week to be Sunday, and embed the time series as a 
$7$-variate weekly time series via a call to **sigex.daily2weekly**.

```{r}
first.day <- 1
data.ts <- sigex.daily2weekly(dataONE.ts,first.day,start.date)
plot(data.ts)
```

- As a result $N= 7$ and $T = 240$, which yields an initial NA (because the 
first data value is on Monday, but the week begins on Sunday). Also the data 
ends on a Sunday, so the last week has six NA values.
- This is a classic example of a *ragged edge* data pattern.

##  Model Declaration

- Now we want to define our time series model.
- We begin by defining dimension $N$ and sample size $T$.
- Note that although the data analyzed is univariate, because of the embedding 
now $N=7$.

```{r}
N <- dim(data.ts)[2]
T <- dim(data.ts)[1]
```

- Next, we define appropriate subspans of the holiday regressors.
- The *easter* regressor is based on the Easter day itself and also the day 
before Easter.
- The *school1* regressor corresponds to the beginning of the first school 
holiday, with a window for the day of and the day after.
- The *school2* and *school3* regressors are analogous for the second and third
holidays.
- The *school1e* regressor corresponds to the end of the first school holiday, 
with a window for the day of and the day before.
- The *school2e* and *school3e* regressors  are analogous for the second and 
third holidays.

```{r}
times <- seq((range[[1]][1]-begin[1])*period+(range[[1]][2]-begin[2])+1,
             (range[[2]][1]-begin[1])*period+(range[[2]][2]-begin[2])+1,1)
easter.reg <- NZregs[times,1]
school1.reg <- NZregs[times,2]
school1e.reg <- NZregs[times,3]
school2.reg <- NZregs[times,4]
school2e.reg <- NZregs[times,5]
school3.reg <- NZregs[times,6]
school3e.reg <- NZregs[times,7]
```

- These are the regressors for daily data, and need to be embedded as weekly 
series.

```{r}
easter.reg <- sigex.daily2weekly(easter.reg,first.day,start.date)
school1.reg <- sigex.daily2weekly(school1.reg,first.day,start.date)
school1e.reg <- sigex.daily2weekly(school1e.reg,first.day,start.date)
school2.reg <- sigex.daily2weekly(school2.reg,first.day,start.date)
school2e.reg <- sigex.daily2weekly(school2e.reg,first.day,start.date)
school3.reg <- sigex.daily2weekly(school3.reg,first.day,start.date)
school3e.reg <- sigex.daily2weekly(school3e.reg,first.day,start.date)
```

- Next, we replace any ragged edge NA values (arising from embedding) with zero, 
which is appropriate because the corresponding data values are also missing.  

```{r}
easter.reg[is.na(easter.reg)] <- 0
school1.reg[is.na(school1.reg)] <- 0
school1e.reg[is.na(school1e.reg)] <- 0
school2.reg[is.na(school2.reg)] <- 0
school2e.reg[is.na(school2e.reg)] <- 0
school3.reg[is.na(school3.reg)] <- 0
school3e.reg[is.na(school3e.reg)] <- 0
```

## Model Specification

- The data has a slight upwards trend drift, and there is strong annual 
correlation present.  
- We will fit a SVARMA model for data differenced to stationarity by applying 
annual differencing $1 - L^{52}$. (Here $L$ is the weekly lag operator, and 
there are approximately 52 weeks per year.)  
- We set the model order to a nonseasonal $q=1$ and seasonal $P = 1$.
    
```{r}
mdl <- NULL
mdl <- sigex.add(mdl,seq(1,N),"svarma",c(0,1,1,0,52),NULL,"process",
                 c(1,rep(0,51),-1))
mdl <- sigex.meaninit(mdl,data.ts,0)
```

- The call to **sigex.add** indicates there is one main component.
- The second argument indicates the rank configuration for the process; setting 
*vrank* equal to seq(1,N) ensures a full rank covariance matrix.
- The third argument gives the model type as *svarma*; the order $q=1$, $P=1$ of
the SVARMA is given in the fourth argument.  
- The fifth argument provides parameter bounds, and does not apply to this type 
of model, and hence is set to NULL.
- The sixth argument associates a label.
- The last argument is for the scalar differencing operator $\delta^{(1)} (z)$, 
expressed as a vector of coefficients.
- The specification of the regressors takes more coding, as we loop over the 
days of the week.

```{r}
for(i in 1:N) {
  mdl <- sigex.reg(mdl,i,ts(as.matrix(easter.reg[,i]),
                            start=start(easter.reg),
                            frequency=frequency(easter.reg),
                            names="Easter-day"))
  mdl <- sigex.reg(mdl,i,ts(as.matrix(school1.reg[,i]),
                            start=start(school1.reg),
                            frequency=frequency(school1.reg),
                            names="School1-Start"))
  mdl <- sigex.reg(mdl,i,ts(as.matrix(school1e.reg[,i]),
                            start=start(school1e.reg),
                            frequency=frequency(school1e.reg),
                            names="School1-End"))
  mdl <- sigex.reg(mdl,i,ts(as.matrix(school2.reg[,i]),
                            start=start(school2.reg),
                            frequency=frequency(school2.reg),
                            names="School2-Start"))
  mdl <- sigex.reg(mdl,i,ts(as.matrix(school2e.reg[,i]),
                            start=start(school2e.reg),
                            frequency=frequency(school2e.reg),
                            names="School2-End"))
  mdl <- sigex.reg(mdl,i,ts(as.matrix(school3.reg[,i]),
                            start=start(school3.reg),
                            frequency=frequency(school3.reg),
                            names="School3-Start"))
  mdl <- sigex.reg(mdl,i,ts(as.matrix(school3e.reg[,i]),
                            start=start(school3e.reg),
                            frequency=frequency(school3e.reg),
                            names="School3-End"))
}
```

## Model Estimation

- Next we consider model fitting. 
- Here we make use of parameter constraints for the regressors: because the same 
holiday effect is present for each day of the week (recall that the weekly 
regressors are derived from a single daily regressor), the parameter estimates 
should be constrained to be identical.
- We use the **sigex.constrainreg** function to do this: the *regindex* argument 
specifies indices $J_1, J_2, \ldots, J_N$ that delineate which regressors are 
being considered for each series.
- For instance, $J_1 = \{ 1, 3, 4 \}$  would correspond to the *trend*, 
*school1*, and *school1e* regressors for the first (Sunday) series.
- The *combos* argument provides the linear combinations of these regressors, 
but when set to NULL instead it is enforced that all the regression parameters 
are the same.

```{r}
constraint <- NULL
constraint <- rbind(constraint,sigex.constrainreg(mdl,data.ts,
                                                  list(2,2,2,2,2,2,2),NULL))
constraint <- rbind(constraint,sigex.constrainreg(mdl,data.ts,
                                                  list(3,3,3,3,3,3,3),NULL))
constraint <- rbind(constraint,sigex.constrainreg(mdl,data.ts,
                                                  list(4,4,4,4,4,4,4),NULL))
constraint <- rbind(constraint,sigex.constrainreg(mdl,data.ts,
                                                  list(5,5,5,5,5,5,5),NULL))
constraint <- rbind(constraint,sigex.constrainreg(mdl,data.ts,
                                                  list(6,6,6,6,6,6,6),NULL))
constraint <- rbind(constraint,sigex.constrainreg(mdl,data.ts,
                                                  list(7,7,7,7,7,7,7),NULL))
```

- So the first call here identifies in *regindex* the *easter* regressor for
each of the seven series, and *combos* set to NULL indicates that these 
parameters will be equal.
- The next call does the same for the *school1* regressor, and so forth; the 
first regressor is the *trend*, and this remains unconstrained.   
- The output of **sigex.constrainreg** is a *constraint* matrix, which can be
then passed into the MLE fitting routine. 
    
```{r}    
par.mle <- sigex.default(mdl,data.ts,constraint)
psi.mle <- sigex.par2psi(par.mle,mdl)
```    
    
- Having initialized the pre-parameter vector, we can start nonlinear
optimization.
- In this case the nonlinear optimization can take a long time. So we just read
the final MLE for *psi* in below.

```{r}
psi.mle <- c(0.195543136661102, 0.210896311604314, 0.224884839193806, 
             0.297710366252992, 0.318076964322756, 0.288167112704856, 
             0.130848490409544, 0.17299456406149, 0.213720675984967, 
             0.0780486278573135, 0.188624612762698, 0.467926733192453,
             0.434759495960342, 0.314822770386916, 0.280937951321511, 
             0.484722644882768, 0.413444260303994, 0.193778981158949,
             0.550345260075344, 0.621238984502903, 0.563243057038342,
             -3.99208014979805, -4.19642305454813, -3.8415818035668,
             -4.32856255035426, -3.95499509133263, -3.96279211490641, 
             -4.03911188215587, -0.369778097197869, -0.0318374973193093, 
             -0.0459307709775452, -0.138903858423006, -0.10135335156786,
             -0.2142068950075, -0.112796685064521, -0.140796608637905,
             -0.237750837664742, -0.194243215055276, 0.0219933280900727,
             -0.0185736546749142, 0.139055815995136, 0.185830059881372, 
             0.0297317536468041, -0.0155812148508496, -0.0671396386183212, 
             0.0476284405203113, -0.0570660189385589, -0.155367001953893, 
             -0.198330717820409, -0.0295063266207021, -0.073765478114995, 
             -0.0829227752708303, -0.187804918878073, -0.0422869655675992, 
             -0.0251387557650662, 0.0656914410289482, -0.0840920843255883, 
             -0.145935243196311, -0.0687523983229335, -0.117014892923535, 
             -0.144653364980642, -0.0790943505687107, 0.0611829711400193, 
             -0.104109662533708, 0.112359996172669, -0.124553026055195, 
             -0.163431409482439, -0.0283323592215797, -0.0312605011210402, 
             -0.178946106450384, -0.266581746889425, -0.164720047360709, 
             0.00959770743708815, 0.159928359498417, -0.0132411806278071,
             -0.0642101491748187, -0.143487318464752, -0.086691994085389,
             0.0814467471639927, -0.0123455935133012, 0.00908782055333108,
             -0.0441428053848971, -0.0194126102991047, -0.00524075702955244,
             -0.0105279430631441, -0.00420628604690744, 0.0789699162821441,
             -0.015850906929479, -0.043499933859938, -0.0062803013365077,
             -0.0207414587537556, -0.0459309858718756, 0.0445914128582114,
             -0.0946876132299148, 0.00588335338136514, 0.00922656520387984,
             -0.0140866870525241, -0.0157842160720905, -0.0319375488373468,
             0.0514126088016708, 0.0141484170525309, -0.124101882618575, 
             -0.00774709495827467, 0.0255995670388491, -0.0364850207049338, 
             0.0230572817360244, 0.0220446070005591, -0.000478476269822034, 
             -0.0168616246820038, -0.102427194747844, 0.00112763406309242, 
             -0.0259316390382219, -0.010758329862067, -0.00198961225224573, 
             -0.0464823007621206, -0.0165414789720724, 0.0104091697972505, 
             -0.14827440476309, 0.0505648610397286, -0.0183900737977979, 
             0.0311301005304512, 0.00518380184796121, 0.0104592851480601,
             -0.0109208667318225, -0.0637859237889965, -0.0656126331052389, 
             0.00033010004554818, -0.191533893764455, -0.157111537080312, 
             0.141669136870683, -0.0225792281088362, 0.157100658532765,
             -0.08441684545665, 0.1847632577291, 0.000561604448005, 
             -0.191533893764455, -0.157111537080311, 0.141669136870683, 
             -0.0225792281088361, 0.157100658532765, -0.0844168454566499,
             0.183411740706148, 0.000312708939566006, -0.191533893764455,
             -0.157111537080312, 0.141669136870683, -0.0225792281088362,
             0.157100658532765, -0.08441684545665, -0.00163601810705088, 
             0.00100594355939265, -0.191533893764455, -0.157111537080311,
             0.141669136870683, -0.0225792281088362, 0.157100658532765,
             -0.08441684545665, 0.0156601320968592, 9.68417444876468e-05,
             -0.191533893764455, -0.157111537080312, 0.141669136870683, 
             -0.0225792281088361, 0.157100658532765, -0.08441684545665,
             0.00328023110252021, 0.000461006050610564, -0.191533893764455, 
             -0.157111537080312, 0.141669136870683, -0.0225792281088361,
             0.157100658532765, -0.08441684545665, 0.00430421751741802,
             0.000496876409071807, -0.191533893764455, -0.157111537080311, 
             0.141669136870683, -0.0225792281088362, 0.157100658532765,
             -0.08441684545665, 0.0023532873975757)
par.mle <- sigex.psi2par(psi.mle,mdl,data.ts)
```

- We obtain, format, and store the residuals.
- Then we compute the autocorrelations, and plot.

```{r}
resid.mle <- sigex.resid(psi.mle,mdl,data.ts)[[1]]
resid.mle <- sigex.load(t(resid.mle),start(data.ts),frequency(data.ts),
                        colnames(data.ts),TRUE)
resid.acf <- acf(resid.mle,lag.max=4*53,plot=FALSE)$acf
par(mfrow=c(N,N),mar=c(3,2,2,0)+0.1,cex.lab=.8,cex.axis=.5,bty="n")
for(j in 1:N)
{
  for(k in 1:N)
  {
    plot.ts(resid.acf[,j,k],ylab="",xlab="Lag",ylim=c(-1,1),cex=.5)
    abline(h=1.96/sqrt(T),lty=3)
    abline(h=-1.96/sqrt(T),lty=3)
  }
}
```

- We can also examine the condition numbers, to see if there is rank reduction
possible.
- Typically there won't be any rank reduction with a single component model.

```{r}
log(sigex.conditions(data.ts,psi.mle,mdl))
```

- We can examine the Portmanteau statistic for residual serial correlation. 
- We use 2 years of lags.

```{r}
sigex.portmanteau(resid.mle,2*52,length(psi.mle))
```

- The output indicates there is moderate  residual autocorrelation, which is 
confirmed by inspection of the sample autocorrelation.
- We can also inspect normality via the Shapiro-Wilks normality test.

```{r}
sigex.gausscheck(resid.mle)
```

- Having completed our model analysis, we store our results in a single list
object.

```{r} 
analysis.mle <- sigex.bundle(data.ts,transform,mdl,psi.mle)
```


## Signal Extraction

- Now we consider applying a daily seasonal adjustment filter that we embed as a
$7$-variate weekly filter.  
- First we load up the fitted model for signal extraction.

```{r}
data.ts <- analysis.mle[[1]]
mdl <- analysis.mle[[3]]
psi <- analysis.mle[[4]]
param <- sigex.psi2par(psi,mdl,data.ts)
```

- The daily seasonal adjustment filter *sa.hifilter* is an extension of the 
classic crude trend filter of X11 from monthly to daily frequency.
- It should suppress all daily frequencies of the form $2\pi j/365$ for 
$1 \leq j \leq 183$, while also passing linear trends. This is embedded with a 
call to **sigex.hi2low**, where the high and low frequencies are indicated and 
the *shift.hi* parameter is $c = 183$, corresponding to a symmetric filter.
- The output is the $7$-variate weekly filter *sa.lowfilter*, with shift 
parameter $C = 27$ given by *shift.low*.

```{r}
sa.hifilter <- c(1,rep(2,365),1)/(2*365)
len <- 183
hi.freq <- 7
low.freq <- 1
shift.hi <- len
out <- sigex.hi2low(sa.hifilter,hi.freq,low.freq,shift.hi)
sa.lowfilter <- out[[1]]
shift.low <- out[[2]]
```

- This filter is applied using a standard call of **sigex.adhocextract**.
- This function automatically supplies missing value imputations for the ragged 
edges, and does forecast and aftcast extension for application of the filter. 
However, this should be transformed back to a daily series, so that we can view 
the daily series with its seasonal adjustment.
- The function  **sigex.weekly2daily** transforms the series, using the
indicated value of *first.day*.
- We repeat this for the lower and upper values of the confidence intervals.       

```{r}
sa.low <- sigex.adhocextract(psi,mdl,data.ts,sa.lowfilter,shift.low,0,TRUE)
sa.hi.daily <- list()
sa.hi.daily[[1]] <- sigex.weekly2daily(ts(sa.low[[1]],start=start(dataONE.ts),
                                frequency=frequency(dataONE.ts)),first.day)
sa.hi.daily[[2]] <- sigex.weekly2daily(ts(sa.low[[2]],start=start(dataONE.ts),
                                frequency=frequency(dataONE.ts)),first.day)
sa.hi.daily[[3]] <- sigex.weekly2daily(ts(sa.low[[3]],start=start(dataONE.ts),
                                frequency=frequency(dataONE.ts)),first.day)
```

- For comparison we can design another filter that only removes the weekly 
dynamics.  
- This is done by considering a simple filter annihilating weekly effects from 
the daily data, namely $(1 + B + \ldots + B^6)/7$, which we call the Trading 
Day (TD) filter.  

```{r}
td.hifilter <- rep(1,7)/7
len <- 3
hi.freq <- 7
low.freq <- 1
shift.hi <- len
out <- sigex.hi2low(td.hifilter,hi.freq,low.freq,shift.hi)
td.lowfilter <- out[[1]]
shift.low <- out[[2]]
```

- Now $c = 3$ and $C = 1$ for the TD filter.   

```{r}
td.low <- sigex.adhocextract(psi,mdl,data.ts,td.lowfilter,shift.low,0,TRUE)
td.hi.daily <- list()
td.hi.daily[[1]] <- sigex.weekly2daily(ts(td.low[[1]],start=start(dataONE.ts),
                                frequency=frequency(dataONE.ts)),first.day)
td.hi.daily[[2]] <- sigex.weekly2daily(ts(td.low[[2]],start=start(dataONE.ts),
                                frequency=frequency(dataONE.ts)),first.day)
td.hi.daily[[3]] <- sigex.weekly2daily(ts(td.low[[3]],start=start(dataONE.ts),
                                frequency=frequency(dataONE.ts)),first.day)
```

- Before plotting, the trend must be computed. 
- A subtlety is that we have $7$ trends, one for each day of week, which 
therefore combine long-term trend effects and day-of-week effects. The mean of 
the $7$ parameter values corresponds to the overall mean of each week, and the 
residual is the day-of-week effect in excess of the mean.   
- First *reg.td* is computed, which is the trend means for each day-of-week 
expressed as a daily time series. We augment by one prior week's values, which 
would be all zero, and then convert to a daily series.
- Next, *reg.trend* is obtained by filtering with $(1 + B + \ldots + B^6)/7$, 
and finally remove the NA value (resulting from using **filter**, which does not 
extend the input series).  
 
```{r}
reg.td <- NULL
for(i in 1:N){reg.td <- cbind(reg.td,sigex.fixed(data.ts,mdl,i,param,"Trend"))}
reg.td <- rbind(rep(0,N),reg.td)
reg.td <- ts(sigex.weekly2daily(reg.td,first.day),
                start=start(dataONE.ts),frequency=period)
reg.trend <- stats::filter(reg.td,td.hifilter,method="convolution",sides=1)
reg.trend <- as.matrix(reg.trend[8:length(reg.td)])
```

- The illustration is completed with a display: the seasonal adjustment contains 
some mild weekly effects (because the filter is designed to remove daily, and 
not weekly frequencies), whereas the TD-adjusted component has annual movements 
but not weekly effects.

```{r}
trendcol <- "tomato"
cyccol <- "orchid"
seascol <- "seagreen"
sacol <- "navyblue"
fade <- 60
plot(dataONE.ts,xlab="Year")
sigex.graph(sa.hi.daily,reg.trend,start(sa.hi.daily[[1]]),
            period,1,0,trendcol,fade)
sigex.graph(td.hi.daily,reg.trend,start(td.hi.daily[[1]]),
            period,1,0,sacol,fade)
```


